# Introduction
A regression model is a way to express a relationship between some response and one or more predictors. In order to estimate the model, you need to have measurements (aka observations) of the response and the predictors. We may use the following terms interchangeably:

- response
- response variable
- dependent variable
- output
- output variable

Likewise, the following may be used interchangeably:

- predictor
- predictor variable
- independent variable
- input
- input variable


## Outline
These are the parts of the workshop:

### Part one
1. Linear model
2. Categorical vs. continuous features
3. Generalized linear model

### Part two
4. Random effects nd mixed-effects models

--------------------
## The linear model

A **linear model** is an equation that describes some output variable as a linear function of some input variable(s).

Note that the linear model is a relationship, while **linear regression** is a method of estimating that relationship. But since it is by far the most common way of estimating that relationship, the terms are often used interchangeably.

The data for a linear model are typically in a tabular format (imagine a
spreadsheet), where each row of data is called an **observation** and each
column is called a **feature**. The column that is the outcome of the model is
called the **response**. Each observation should include a value for every
feature (there are some ways of handling missing data but that's beyond our
scope for this workshop).

### Preparation

Your computer will do whatever you tell it to do, even if it's not a good idea. With this great power comes the responsibility to think, and to check your assumptions.

The first one to mention is the assumption that there is a relationship between the features and the response, of the type that the model describes. Your first, best way to test that assumption is to plot the data. Summaries like the means, variances, and correlations can only tell you so much. The following example points out why.

The Datasaurus Dozen are a collection of thirteen data sets. Each consists of two features (x and y) that have the same means, variances, and correlations.



```{r install-datasaurus}
#| include: false
#| eval: false
#| echo: false
remotes::install_github("jumpingrivers/datasauRus")
```


```{r plot-datasaurus}
#| message: false
#| warning: false
#| echo: false

library(datasauRus)
library(dplyr)
library(ggplot2)
library(tidyr)

group_by(datasaurus_dozen, dataset) |>
  summarize(correlation = round(cor(x, y), 3), across(c('x', 'y'), .fns=list('mn'=~round(mean(.x), 1), 'var'= ~ round(var(.x), 1))))

ggplot(datasaurus_dozen, aes(x = x, y = y, colour = dataset)) +
    geom_point() +
    theme_void() +
    theme(legend.position = "none") +
    facet_wrap(~dataset, ncol = 3)
```


You probably wouldn't use a linear model for most of the panels of that plot because there isn't a linear relationship between the feature (x-direction) and the response (y-direction). Plotting the data can also reveal problems or oddities in the data that will guide your further investigation.
